---
title: "SVD"
author: "Kachun Huang"
date: "2025-03-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SVD in R

## Generalized Inverse

Consider the following system of linear equations:

$$
\begin{pmatrix}
2 & 1 \\ 4 &2
\end{pmatrix} 
\begin{pmatrix}
y_1 \\ y_2
\end{pmatrix} =
\begin{pmatrix}
1 \\ 1
\end{pmatrix} 
$$

Since

$$
\begin{pmatrix}
2 & 1 \\ 4 &2
\end{pmatrix}
$$

is a Singular matrix, where the determinant is 0. So there is no exact solution but it is possible to get a solution in the least-squares sense.

Recall, the definition of inverse, given $A$ is non-singular:

$$
AA^{-1}A = A
$$

Now suppose $A$ is singular, then the inverse $A^{-1}$ doesn't exist.

But, we can compute a Moore-Penrose generalized Inverse $A^+$, such that

$$
AA^+A = A
$$

where $A^+$ is given by

$$
A^+ = (A^TA)^{-1}A^T
$$

Relation to SVD, Moore-Penrose generalized Inverse $A^+$ is given by

If $A = U\Sigma V^T$, then $A^+ = V\Sigma^+ U^T$

where,

-   $U$ is a $n \times p$ orthonormal matrix

-   $V$ is a $p \times p$ orthonormal matrix

-   $\Sigma$ is the $p \times p$ diagonal matrix

-   $\Sigma^+$ is the $\Sigma$ diagonal matrix but take reciporical of all diagonal entries and they have the same rank\

```{r}
X <- matrix(
  c(2,4,1,2),
  nrow=2
)
X
```

Do SVD of X

```{r}
svd_X <- svd(X)
svd_X
```

```{r}
U <- svd_X$u
Sigma <-diag(svd_X$d)
V <- svd_X$v
```

Find $\Sigma^+$

```{r}
Sigma_plus <- diag(c(1/svd_X$d[1],0))
Sigma_plus
```

Find $X^+$

```{r}
Xinv <- V %*% Sigma_plus %*% t(U)
Xinv
```

We can also use `ginv()` from MASS Library

```{r}
library(MASS)
ginv(X)
```

So with the $A^+$, we can compute the solution of this system of linear equation

$$
x = A^+y
$$

```{r}
Xinv %*% c(1,1)
```

## Least Square Estimation

Consider the multiple linear regression model in matrix notation:

$$
Y = \beta X
$$

Then we can derive estimation of the $\beta$ using Moore-Penrose generalized Inverse

$$
\hat{\beta} = (X^TX)^{-1}X^TY
$$

Example

```{r}
set.seed(123)
n <- 100
x1 <- rnorm(n)
x2 <- rexp(n)
y <- 2-1*x1+3*x2+rnorm(n)

X <- cbind(1,x1,x2)
beta.est <- solve(t(X) %*% X) %*% t(X) %*% y
beta.est
```

OLS summary

```{r}
fit <- lm(y ~ X)
summary(fit)
```
