---
title: "Neural Network Regression"
author: "Kachun Huang"
date: "2025-03-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Neural Network Regression

## library Import

```{r}
library(neuralnet)
library(tidyverse)
library(caret)
library(MASS)
library(knitr)
```

## Data Import

```{r}
data("Boston")
house = Boston
head(house)
```

```{r}
str(house)
```

The Boston housing price data we will use for our homework is taken from the Kaggle\
competition site ([https://www.kaggle.com/vikrishnan/boston-house-prices).](https://www.kaggle.com/vikrishnan/boston-house-prices).) We will treat this dataset as our entire data.

\
Each record in the database describes a Boston suburb or town. The data was drawn from the Boston Standard Metropolitan Statistical Area (SMSA) in 1970.

Our data has 506 observations and 14 variables:

\
Attribute Information (in order):

-   CRIM: per capita crime rate by town

-   ZN: proportion of residential land zoned for lots over 25,000 sq.ft.

-   INDUS: proportion of non-retail business acres per town

-   CHAS: Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)

-   NOX: nitric oxides concentration (parts per 10 million)

-   RM: average number of rooms per dwelling

-   AGE: proportion of owner-occupied units built before 1940

-   DIS: weighted distances to five Boston employment centers

-   RAD: index of accessibility to radial highways

-   TAX: full-value property-tax rate per 10,000 dollars

-   PTRATIO: pupil-teacher ratio by town

-   Black: The result of the equation B=1000(Bk - 0.63)\^2 where Bk is the\
    proportion of blacks by town

-   LSTAT: %lower status of the population

-   MEDV: Median value of owner-occupied homes in 1000 dollars

## Data Pre-processing

### Exclude RAD Columns

```{r}
house <- subset(house, select = -c(rad))
head(house)
```

```{r}
str(house)
```

### Compute Mean and sd (for descaling)

```{r}
mean <- mean(house$medv)
sd <- sd(house$medv)
```

## Normalize Data

```{r}
house <- data.frame(scale(house))
head(house)
```

## Data Splitting

```{r}
set.seed(123)
training.sample <- house$medv %>% createDataPartition(p=0.75, list=FALSE)
train.data <- house[training.sample,]
test.data <- house[-training.sample,]
```

In your own research, you can follow this:

1)  Always scale continuous features before fitting a neural network.

2)  Categorical variables (one-hot encoded) do not need scaling.

## Neural Network with no hidden layer

```{r}
nn_no_model <- neuralnet(
  medv ~.,
  train.data,
  hidden = 0,
  err.fct = "sse",
  linear.output = TRUE
)
```

### Plot

```{r}
plot(nn_no_model, rep="best")
```

### Predict on Test Data

```{r}
#predict on test data
pred_nn_no <- predict(nn_no_model, test.data)
```

### Descale

```{r}
# compute y(unscaled)
y <- test.data$medv * sd + mean
#compute y^hat (unscaled)
yhat <- pred_nn_no*sd+mean
```

### Compute Test MSE

```{r}
(MSE.nn.1 = RMSE(y,yhat)^2)
```

or

```{r}
sum((y-yhat)^2)/length(y)
```

## Neural Network with 1 hidden layer

```{r}
nn_one_model <- neuralnet(
  medv ~.,
  train.data,
  hidden = 3,
  err.fct = "sse",
  linear.output = TRUE
)
```

### Plot

```{r}
plot(nn_one_model, rep="best")
```

### Predict on Test Data

```{r}
#predict on test data
pred_nn_one <- predict(nn_one_model, test.data)
```

### Descale

```{r}
# compute y(unscaled)
y <- test.data$medv*sd+mean

#compute y^hat (unscaled)
yhat <- pred_nn_one*sd+mean
```

### Compute Test MSE

```{r}
(MSE.nn.2 = RMSE(y,yhat)^2)
```

or

```{r}
sum((y-yhat)^2)/length(y)
```

## Multiple Linear Regression

```{r}
mlr <- lm(medv~., data=train.data)
```

### Summary

```{r}
summary(mlr)
```

### Predict on Test Data

```{r}
pred_mlr <- predict(mlr, test.data)
```

### Descale

```{r}
# compute y(unscaled)
y <- test.data$medv*sd+mean

#compute y^hat (unscaled)
yhat <- pred_mlr*sd+mean
```

### Compute Test MSE

```{r}
(MSE.mlr = RMSE(y,yhat)^2)
```

or

```{r}
sum((y-yhat)^2)/length(y)
```

## Compare Test MSE of all models

```{r}
print(paste(MSE.nn.1, MSE.nn.2, MSE.mlr))
```

```{r}
final1 <- data.frame(predictions_NN0=pred_nn_no*sd+mean, predictions_NN1=pred_nn_one*sd+mean,predictions_MLR = pred_mlr*sd+mean, actual_repsonse = test.data$medv*sd+mean)
kable(head(final1))
```

## Plot predicted vs. observed

```{r}
plot(
  test.data$medv*sd+mean, 
  pred_nn_no*sd+mean,
  col="red",
  ylab = "predicted house price",
  xlab= "true values of house price",
  )

# plot nn with 1 layer 3 neurons
points(test.data$medv*sd+mean, pred_nn_one*sd+mean, col="blue")

#plot mlr 
points(test.data$medv*sd+mean, pred_mlr*sd+mean, col = "green")
abline (a=0, b=1)
```

```{r}
plot(
  pred_mlr,
  pred_nn_no,
  col = "blue",
  ylab = "prediction of NN with no hidden layer",
  xlab = "prediction of MLR"
)
abline(a=0, b =1)
```
