---
title: "Monte Carlo Integration"
author: "Kachun Huang"
date: "2025-04-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Monte Carlo Integration

Monte Carlo Integration is a statistical method based on random sampling

Idea:

-   If $X$ is a random variable with pdf $f(x)$, then

$$
E[g(X)] = \int_{-\infty}^{\infty}g(x)f(x)dx
$$

-   Thus, to estimate $\omega = \int_a^bg(x)dx$, we do

$$
X_1,...,X_n \sim U(a,b) \text{, then } \hat{\omega} = \frac{1}{n}\sum_{i=1}^{n}g(X_i)
$$

## Algorithm

1) Generate iid $X_1,...,X_n$ from $U(a,b)$

2) Compute $\bar{g(X)} = \frac{1}{n}\sum_{i=1}^{n}g(X_i)$

3) then $\hat{\omega} = (b-a)\bar{g(X)}$

## Example

Compute a Monte Carlo Estimate of $\omega = \int_0^1e^{-x}dx$

```{r}
set.seed(123)
n<-10000

x <- runif(n, min=0, max=1)
f_x <- exp(-x)

#Average of g(X)
omega_hat <- mean(f_x)

#compute (b-a)mean(g(X))
(1-0)*omega_hat
```

Compute a Monte Carlo Estimate of $\omega = \int_2^4e^{-x}dx$

```{r}
set.seed(123)
n<-10000

x <- runif(n, min=2, max=4)
f_x <- exp(-x)

#Average of g(X)
omega_hat <- mean(f_x)

#compute (b-a)mean(g(X))
(4-2)*omega_hat
```

Estimate $\phi(-1.645) = \int_{-\infty}^{-1.645} \frac{1}{\sqrt(2\pi)}e^{\frac{-t^2}{2}}dt$ using Monte Carlo Method

-   Generating $n$ random samples $X_1,...,X_n$ from a standard normal distribution $N(0,1)$

-   Counting how many samples satisfy $X_i \le -1.645$

-   Computing the proportion:

$$
\phi(-1.645) = \frac{\# \text{ of } X_i \le -1.645}{n}
$$

```{r}
set.seed(123)
n<-10000

samples <- rnorm(n)

count <- sum(samples <= -1.645)

estimate <- count/n 
estimate # or estimate <- mean(samples <= -1.645)
```

# Importance Sampling

Suppose we want to estimate:

$$
\hat{\mu}=E[g(X)]=\int g(x)f(x)dx,
$$

where $X \sim f(x)$, and $f(x)$ is the density of $X$. We can rewrite the expectation using another density $h(x)$ (the importance distribution) as long as $h(x) > 0$whenever $h(x)f(x) \ne 0$ (to avoid division by zero):

$$
\mu=\int h(x)f(x)dx=\int g(x) \frac{f(x)}{h(x)}h(x) dx= E_h[g(x)\frac{f(x)}{h(x)}]
$$

where the expectation is now taken over $X \sim h(x)$

## Algorithm

The importance sampling estimator is:

1.  Draw $n$ samples $X_1,...,X_n \sim h(x)$

2.  Compute the weights $\omega(X_i) = \frac{f(X_i)}{h(X_i)}$

3.  Estimate $\mu$ as:

$$
\mu = \frac{1}{n}\sum_{i=1}^{n}g(X_i)\frac{f(X_i)}{h(X_i)}
$$

## Example

Simulate $\phi(-2) = P(Z \le -2)$ using importance sampling with $h(x) \sim Cauchy(1)$

```{r}
set.seed(123)
n<-10000

x <- rcauchy(n)

f <- function(x){
  return(dnorm(x))
}

h <- function(x){
  return(dcauchy(x))
}

g <- function(x){
  return(x <= -2)
}

estimate <- mean(g(x) * (f(x)/h(x)))
estimate
```
