---
title: "CART in R"
author: "Kachun Huang"
date: "2025-03-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# CART in R

## Import Dependencies

```{r}
library(rpart)
library(rpart.plot)
library(rattle)
library(caTools)
library(tidyverse)
library(caret)
```

## Model Fitting

syntax:

```{r eval=FALSE, include=TRUE}
model <- rpart(
  Y~.,
  data, 
  method,
  control = rpart.control(cp=...)
)
```

parameters:

-   Method: one of **`"anova"`**, **`"poisson"`**, **`"class"`** or **`"exp"`**. If **`method`** is missing then the routine tries to make an intelligent guess. If **`y`** is a survival object, then **`method = "exp"`** is assumed, if **`y`** has 2 columns then **`method = "poisson"`** is assumed, if **`y`** is a factor then **`method = "class"`** is assumed, otherwise **`method = "anova"`** is assumed. It is wisest to specify the method directly, especially as more criteria may added to the function in future.

-   control: a list of options that control details of the **`rpart`** algorithm.

    -   in `rpart.control`, `cp` is the complexity parameter.

        -   `cp` controls the trade-off between tree size and accuracy

        -   when `cp` = 0, creates a fully grown tree. more splits, more details but overfitting

### Fit Controlled Model

To avoid overfitting, we can set constraints like `minsplit` and `minbucket`.

-   `minsplit`: Minimum number of observations required to split a node.

-   `minbucket`: Minimum number of observations in a leaf node.

```{r eval=FALSE, include=TRUE}
model <- rpart(
  Y~.,
  data, 
  method,
  control = rpart.control(cp=..., minsplit=..., minbucket =...)
)
```

### Fit a Tree Using Entropy

By default, `rpart` uses the Gini index for splits. To use entropy instead, specify the `split` parameter.

```{r eval=FALSE, include=TRUE}
model <- rpart(
  Y~.,
  data,
  method,
  params = list(split="information") #information gain (entropy)
)
```

## Model Plot

This is from `rattle` packages

```{r eval=FALSE, include=TRUE}
# Visualize the tree using fancyRpartPlot from rattle
par(xpd = NA)  # Allow plotting outside the margins
fancyRpartPlot(model, 
               sub = "", 
               caption = "", 
               main = "Title_name", 
               split.col = "blue", # Color of split text
               box.palette = "RdYlBu", # Color scheme
               tweak = 1.2 # Adjust node size
              )
```

or we can use `rpart.plot` package

```{r eval=FALSE, include=TRUE}
rpart.plot(
  model, 
  type = 4, # Shows split labels at nodes
  extra = 104, #Displays probabilities and class labels at terminal nodes.
  tweak = 1.2 # Adjusts font size
  )
```

## Prune the tree

Pruning reduces overfitting by selecting an optimal complexity parameter (`cp`) using cross-validation.

```{r eval=FALSE, include=TRUE}
set.seed(123)
model_pruned <- train(
  Y~.,
  data,
  method = "rpart",
  trControl = trainControl("cv", number = 10), # 10 fold CV
  tuneLength = 100 # Tries 100 different complexity parameters (cp) to find the best one.
)
```

To see the Accuracy vs. cp plot

```{r eval=FALSE, include=TRUE}
plot(model_pruned)
```

To access the best `cp`

```{r eval=FALSE, include=TRUE}
model_pruned$bestTune
```

## Prediction

Make predictions on the test set and evaluate performance using a confusion matrix.

```{r eval=FALSE, include=TRUE}
# Predict on test data
pred <- predict(model, test.data, type = "class")

# confusion matrix (caret package)
confusionMatrix(factor(pred), factor(test.data$Response), positive = "1")
```

# Example

## Import Data

```{r}
data <- read.csv("/Users/thomashuang/Downloads/Titanic.csv")
head(data)
```

```{r}
str(data)
```

## Data Pre-processing

### Remove Missing Values

```{r}
data <- na.omit(data)
```

```{r}
str(data)
```

### Factorized categorical variable

```{r}
data$Pclass <- factor(data$Pclass)
data$Survived <- factor(data$Survived)
```

### Select only subset of the features

```{r}
data <- subset(data, select=-c(Name, Ticket, Cabin))
str(data)
```

## Data Splitting

```{r}
set.seed(123)
training.sample <- data$Survived %>% createDataPartition(p=0.8, list = FALSE)
train.data <- data[training.sample,]
test.data <- data[-training.sample,]
```

## Fit Full Tree Model

```{r}
full <- rpart(
  Survived~., 
  data=train.data, 
  control = rpart.control(cp=0) # fully grown tree
  )
```

### Plot

```{r}
fancyRpartPlot(
  full, 
  sub = "", 
  caption = "", 
  main = "Decision Tree for Survived",
  split.col = "blue", # Color of split text
  box.palette = "RdYlBu", # Color scheme
  tweak = 1.2 # Adjust node size
  ) 
```

or use `rpart.plot` pkg

```{r}
rpart.plot(
  full, 
  type = 4, # type = 4: Shows split labels at nodes
  extra = 104, # extra = 104: Displays probabilities and class labels at terminal nodes
  tweak = 1.2, # tweak = 1.2: Adjusts font size
  main = "Decision Tree for Survived"
  )
```

### Prediction

```{r}
pred_full <- predict(
  full,
  test.data,
  type="class"
)
```

### Confusion Matrix

```{r}
confusionMatrix(
  factor(pred_full),
  factor(test.data$Survived),
  positive = "1"
)
```

## Fit Full Tree with Control

```{r}
control_model <- rpart(
  Survived~.,
  data = train.data,
  control = rpart.control(cp=0, 
                          minisplit=10,
                          minbucket=5)
)
```

### Plot

```{r}
fancyRpartPlot(
  control_model, 
  sub = "", 
  caption = "", 
  main = "Decision Tree for Survived",
  split.col = "blue", # Color of split text
  box.palette = "RdYlBu", # Color scheme
  tweak = 1.2 # Adjust node size
)
```

### Prediction

```{r}
pred_control <- predict(
  control_model,
  test.data,
  type="class"
)
```

### Confusion Matrix

```{r}
confusionMatrix(
  factor(pred_control),
  factor(test.data$Survived),
  positive = "1"
)
```

## Fit Tree with (Entropy)

```{r}
model_entropy <- rpart(
  Survived ~ ., 
  data = train.data,
  method = "class", # if resposne is factor
  parms = list(split = "information") # Use entropy
) 
```

### Plot

```{r}
rpart.plot(
  model_entropy, 
  type = 4, 
  extra = 104, 
  tweak = 1.2
)
```

### Prediction

```{r}
pred_entropy <- predict(
  model_entropy,
  test.data,
  type="class"
)
```

### Confusion Matrix

```{r}
confusionMatrix(
  factor(pred_entropy),
  factor(test.data$Survived),
  positive = "1"
)
```

## Pruning full tree with 10-fold CV

```{r}
set.seed(123)

model2 <- train(
  Survived ~., 
  data = train.data, 
  method = "rpart",
  trControl = trainControl("cv", number = 10), #trControl = trainControl("cv", number = 10): Implements 10-fold cross-validation
  tuneLength = 100 # tuneLength = 100: Tries 100 different complexity parameters (cp) to find the best one.
)
```

### Plot

```{r}
plot(model2)
```

The tree starts from its fully grown version, and then cross-validation selects the optimal cp value,\
- 1. The tree is initially grown fully with different cp values.\
- 2. 10-fold cross-validation evaluates tree performance for each cp.\
- 3. The best cp (minimizing validation error) is selected.\
- 4. The final tree is pruned using the best cp.

cp stands for Complexity Parameter:\
- The best cp is the one that minimizes the cross-validation error

### Access the best tune

```{r}
model2$bestTune
```

### Plot the best model

```{r}
fancyRpartPlot(model2$finalModel)
```

### Prediction

```{r}
pred_2 <- predict(
  model2,
  test.data
)
```

### Confusion Matrix

```{r}
confusionMatrix(
  factor(pred_2),
  factor(test.data$Survived),
  positive = "1"
)
```

## Manually Prune the Tree with rpart

```{r}
full_tree <- rpart(Survived ~ ., data = train.data, cp = 0) # Fully grown tree
printcp(full_tree) # Check complexity parameters
```

### Access the best cp

```{r}
best_cp <- full_tree$cptable[which.min(full_tree$cptable[, "xerror"]), "CP"]
best_cp
```

### Plot the best tree

```{r}
pruned_tree <- prune(
  full_tree,
  cp=best_cp
)

rpart.plot(pruned_tree)
```

### Prediction

```{r}
pred_pruned <- predict(
  pruned_tree,
  test.data,
  type = "class"
)
```

### Confusion Matrix

```{r}
confusionMatrix(
  factor(pred_pruned),
  factor(test.data$Survived),
  positive = "1"
)
```

## Result

```{r}
#full tree
table(pred_full, test.data$Survived)
```

```{r}
#pruned tree
table(pred_pruned, test.data$Survived)
```

The pruned model increases TN (79 vs. 71), meaning it is better at correctly identifying dead passengers.

\
The pruned model has a lower TP (36 vs. 47), meaning it predicts fewer survived passengers correctly than the full model.

There is a trade-off. Chose the model based on your research goal.\
