---
title: "Clustering in R"
author: "Kachun Huang"
date: "2025-04-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clustering in R

This tutorial introduces three common clustering methods in R: **Hierarchical Clustering**, **K-means Clustering**, and **Fuzzy C-Means Clustering**, using the `USArrests` dataset. It follows a standard structure: **load libraries → learn syntax → run examples**.

## Load Required Libraries

```{r load-libs, message=FALSE, warning=FALSE}
# Core clustering and plotting libraries
library(stats)    
library(cluster)
library(factoextra)
library(ggplot2)  
```

## Basic Syntax & Functions

### Distance Matrix for Hierarchical Clustering

```{r eval=FALSE, include=TRUE}
d <- dist(data)
```

-   Computes pairwise **Euclidean distance** between observations

### Hierarchical Clustering

```{r eval=FALSE, include=TRUE}
hc <- hclust(d, method = "ward.D2")
```

Performs agglomerative clustering using:

-   `"ward.D2"` – squared distances

    -   other linkage strategies

        -   `ward.D` -Ward’s minimum variance method (classic version)

        -   `single` -Minimum (single) linkage

        -   `complete` -Maximum (complete) linkage

        -   `average` -Average linkage (UPGMA)

        -   `centroid` -Centroid linkage (UPGMC)

#### Plot Dendrogram

```{r eval=FALSE, include=TRUE}
# Dendrogram
plot(hc)

# Add rectangles for k clusters
rect.hclust(hc)
```

or

```{r eval=FALSE, include=TRUE}
# dendrogram from factoextra
fviz_dend(
  hc, 
  k = 4, # num of clusters
)
```

### K-means Clustering

#### Optimal Number of clusters

```{r eval=FALSE, include=TRUE}
fviz_nbclust(
  data, 
  kmeans, 
  method = "gap_stat"
)
```

Choose the k that maximizes the Gap statistic

```{r eval=FALSE, include=TRUE}
km.res <- kmeans(data, centers = 2, nstart = 25)
```

-   `centers`: number of clusters

-   `nstart`: number of initial random configurations

#### Plot Clusters

```{r eval=FALSE, include=TRUE}
fviz_cluster(km.res, 
             data = data,
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal()
)
```

| Argument | Description |
|----|----|
| `fviz_cluster()` | Main function to **plot clustering results** from `kmeans()`, `hclust()`, or `fanny()` |
| `km.res` | The result object from `kmeans()` — contains cluster assignments and centroids |
| `data = data` | The dataset that was clustered — used to calculate PCA for plotting |
| `ellipse.type = "convex"` | Draws **convex hulls** around each cluster to highlight boundaries (other options: `"norm"`, `"t"` for confidence ellipses) |
| `palette = "jco"` | Uses the **"jco" color palette** (from Journal of Clinical Oncology style) for clusters |
| `ggtheme = theme_minimal()` | Applies a **clean and minimal ggplot2 theme** to the plot for aesthetics |

### Fuzzy C-Means Clustering

K-means: Each point belongs to exactly one cluster (0/1)

Fuzzy C-Means: Each point belongs to all clusters to a degree (between 0 and 1)

```{r eval=FALSE, include=TRUE}
res.fanny <- fanny(data, 2)
```

-   Allows **soft membership** (each point belongs to multiple clusters with probabilities)

#### Plot Clusters

```{r eval=FALSE, include=TRUE}
fviz_cluster(
  res.fanny, 
  ellipse.type = "norm", 
  repel = TRUE,
  palette = "jco", 
  ggtheme = theme_minimal(),
  legend = "right"
)
```

| Argument | Description |
|----|----|
| `fviz_cluster()` | Visualizes the clustering result, in this case from a fuzzy clustering object (`res.fanny`) |
| `res.fanny` | Clustering result from `fanny()`, which includes **soft memberships** for each point |
| `ellipse.type = "norm"` | Draws **normal confidence ellipses** (assuming multivariate normal distribution of points) around each cluster |
| `repel = TRUE` | Prevents **text labels from overlapping** by repelling them slightly apart for readability |
| `palette = "jco"` | Uses the **"Journal of Clinical Oncology" (jco)** color palette for cluster coloring |
| `ggtheme = theme_minimal()` | Applies a **clean ggplot2 theme** with minimal gridlines and distractions |
| `legend = "right"` | Places the **cluster legend** on the right side of the plot for clarity |

# Example

## Import Data

```{r}
data <- USArrests
data <- na.omit(data)
data <- scale(data)
data <- as.data.frame(data)  # Ensure it's a data frame
```

## Hierarchical Clustering

### Distance Matrix for Hierarchical Clustering

```{r}
d <- dist(data)
```

### Hierarchical Clustering

```{r}
hc <- hclust(d, method = "ward.D2")
```

### Plot

```{r}
plot(hc, main = "Hierarchical Clustering Dendrogram")

rect.hclust(hc, k = 4, border = "red")
```

or

```{r}
fviz_dend(
  hc,
  k = 4,
  cex = 0.5, # Label size
  k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#D8D900"),
  color_labels_by_k = TRUE, # Color labels by cluster
  rect = TRUE # Draw rectangles around clusters
)
```

## K-means Clustering

### Optimal Number of clusters

```{r}
fviz_nbclust(data, kmeans, method = "gap_stat") 
```

### K-means Clustering

```{r}
km.res <- kmeans(data, centers = 2, nstart = 25)
```

### Plot

```{r}
fviz_cluster(km.res, 
             data = data,
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal()
)
```

## Fuzzy C-means Clustering

### Fitting model and Plot

```{r}
res.fanny <- fanny(data, 2)

fviz_cluster(
  res.fanny, 
  ellipse.type = "norm", 
  repel = TRUE,
  palette = "jco", 
  ggtheme = theme_minimal(),
  legend = "right"
)
```
