---
title: "Multi-Colinearity"
author: "Kachun Huang"
date: "2025-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Library

```{r}
library(corrplot)
library(car)
library(tidyverse)
library(caret)
library(glmnet)
```

## Load Data

```{r}
data <- read.csv("/Users/thomashuang/Downloads/data_factorize_notScaled.csv") 
```

```{r}
str(data)
```

## Exclude Index Column

```{r}
data <- subset(data, select = -c(X))
```

## Convert Column Type

```{r}
# Define the column names as a character vector
numeric_cols <- c("age", "bp", "bgr", "bu", "sc", "sod", "pot", "hemo", "pcv", "wbcc", "rbcc")

# Convert each column to numeric
for (col in numeric_cols) {
  data[[col]] <- as.numeric(data[[col]])
}

# Define the column names as a character vector
categorical_cols <- c("sg", "al", "su", "rbc", "pc", "pcc", "ba", "htn", "dm", "cad", "appet", "pe", "ane", "class")

# Convert each column to factor
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
}
```

```{r}
str(data)
```

# Correlation Analysis

## Correlation Matrix

```{r}
cor_matrix<-cor(
  subset(data, select = -c(sg, al, su, rbc, pc, pcc, ba, htn, dm, cad, appet, pe, ane, class))
)
cor_matrix
```

## Heat Map

```{r}
corrplot(cor_matrix, method = "color", addCoef.col = "black", # Add values in black
         tl.col = "black", tl.srt = 45, # Tilted labels
         diag = FALSE, # Hide diagonal if desired
         number.cex = 0.7 # Adjust text size
) 
```

-   pcv vs. hemo **0.9**

-   hemo vs. rbcc **0.8**

-   rbcc vs. pcv **0.79**

These 3 combo all have really high correlation.

## Explain

Scientifically speaking, they are all directly related to each other.

-   pcv vs. hemo

    -   The more red blood cells you have (i.e. higher PCV), the more hemoglobin you’re likely to have — they rise and fall together.

-   hemo vs. rbcc

    -   More red blood cells, more hemoglobin total (assuming each cell has roughly the same amount).\
        Again.

-   rbcc vs. pcv

    -   Higher RBC count means a greater proportion of blood is occupied by these cells, higher PCV.

These values suggest moderate multicollinearity between PCV and hemoglobin.

# Penalized Regression

## Prepare Data

```{r}
# Prepare data
X <- model.matrix(~ . - 1, data = data %>% subset(select = -class))
y <- data$class
# Split into training and test sets
set.seed(597)
training.samples <- y %>% createDataPartition(p = 0.7, list = FALSE)
X_train <- X[training.samples, ]
X_test <- X[-training.samples, ]
y_train <- y[training.samples]
y_test <- y[-training.samples]
train.data <- data[training.samples,]
test.data <- data[-training.samples,]
```

## Ridge Regression

```{r}
cv_ridge <- cv.glmnet(
  x = X_train,
  y = y_train,
  family = "binomial",
  alpha = 0,          # Ridge regression
  nfolds = 10,        # 10-fold cross-validation
  type.measure = "class"  # Use misclassification error
)
```

### CV plot

```{r}
plot(cv_ridge)
```

### Best Lambda

```{r}
best_lambda <- cv_ridge$lambda.min
cat("Optimal lambda:", best_lambda, "\n")
```

### Model with best lambda

```{r}
ridge_model <- glmnet(
  x = X_train,
  y = y_train,
  family = "binomial",
  alpha = 0,
  lambda = best_lambda
)
```

```{r}
coef(ridge_model)
```

### Prediction

```{r}
pred_prob <- predict(ridge_model, 
                    newx = X_test, 
                    type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
```

### Confusion Matrix

```{r}
cm_ridge <- confusionMatrix(factor(pred_class), y_test, positive = "1")
print(cm_ridge)
```

## Lasso

```{r}
cv_lasso <- cv.glmnet(
  x = X_train,
  y = y_train,
  family = "binomial",
  alpha = 1,          # Lasso regression
  nfolds = 10,        # 10-fold cross-validation
  type.measure = "class"  # Use misclassification error
)
```

### CV plot

```{r}
plot(cv_lasso)
```

### Best Lambda

```{r}
best_lambda <- cv_lasso$lambda.min
cat("Optimal lambda:", best_lambda, "\n")
```

### Model with best lambda

```{r}
lasso_model <- glmnet(
  x = X_train,
  y = y_train,
  family = "binomial",
  alpha = 1,
  lambda = best_lambda
)
```

```{r}
coef(lasso_model)
```

### Prediction

```{r}
pred_prob <- predict(lasso_model, 
                    newx = X_test, 
                    type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
```

### Confusion Matrix

```{r}
cm_lasso <- confusionMatrix(factor(pred_class), y_test, positive = "1")
print(cm_lasso)
```

## Elastic Net

```{r}
# Define cross-validation control
ctrl <- trainControl(
  method = "cv",          # Cross-validation
  number = 10,            # 10 folds
  classProbs = TRUE,      # For probability predictions
  summaryFunction = twoClassSummary,  # For ROC/AUC metrics
  verboseIter = TRUE      # Show progress
)

# response need to be name
train.data$class <- ifelse(train.data$class == "1","ckd","notckd")

set.seed(123)  # For reproducibility
model <- train(
  class~.,              # Predict class using all predictors
  data = train.data,
  method = "glmnet",
  trControl = ctrl,
  metric = "ROC"          # Optimize for AUC
)
```

### Best Tune

```{r}
model$bestTune
```

### Model with bestTune

```{r}
model_en <- glmnet(
  x = X_train,
  y = y_train,
  alpha=model$bestTune$alpha,
  lambda=model$bestTune$lambda,
  family = "binomial"
)
```

```{r}
coef(model_en)
```

### Prediction

```{r}
pred_prob <- predict(model_en, 
                    newx = X_test, 
                    type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
```

### Confusion Matrix

```{r}
cm_en <- confusionMatrix(factor(pred_class), y_test, positive = "1")
print(cm_en)
```

# Dimensionality Reduction

## PCA

```{r}

# Step 2: Perform PCA
pca_result <- prcomp(data, center = TRUE, scale. = TRUE)

# Step 3: Explore results
# Summary of PCA (proportion of variance explained)
summary(pca_result)

# Scree plot to visualize variance explained by each component
#plot(pca_result, type = "lines", main = "Scree Plot")
```
