---
title: "PCA in R"
author: "Kachun Huang"
date: "2025-04-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PCA in R

## Import Library

```{r}
library(factoextra)
library(stats)
library(cluster)
library(factoextra)
library(ggplot2)
library(caret)
library(ggbiplot)
```

## Step 1: Scale

```{r eval=FALSE, include=TRUE}
data_scaled <- scale(data)
```

PCA is affected by the scale of the variables, so standardization is usually necessary.

## Step 2: apply prcomp()

```{r eval=FALSE, include=TRUE}
pca_data <- prcomp(
  data_scaled, 
  center=TRUE, 
  scale.=TRUE # if you didn't scale the data
) 
```

If scaled

```{r eval=FALSE, include=TRUE}
pca_data <- prcomp(
  data_scaled, 
  center=FALSE, 
  scale.=FALSE # if you didn't scale the data
) 
```

### Summary

```{r eval=FALSE, include=TRUE}
summary(pca_data)
```

#### Output includes:

| Statistic | Description |
|----|----|
| **Standard deviation** | The square roots of the eigenvalues of the covariance/correlation matrix. It shows how much variance each principal component (PC) captures in absolute terms. |
| **Proportion of Variance** | The percentage of the total variance in the dataset explained by each PC. |
| **Cumulative Proportion** | The cumulative percentage of total variance explained as you add more PCs. Useful for deciding how many components to keep. |

#### What you can access in the `pca_data` object:

| Component | Description |
|----|----|
| `$sdev` | Standard deviations of the principal components (used to compute variance). |
| `$rotation` | **The matrix of eigenvectors (i.e., loadings).** Columns = PCs, rows = original features. Shows how each variable contributes to each PC. |
| `$center` | The mean of each variable used for centering. |
| `$scale` | The standard deviation used for scaling (if `scale. = TRUE`). |
| `$x` | The **principal component scores**, i.e., the projected data in the new PC space. Each row = observation, each column = PC. |

## Step 3: Scree Plot

We need to use Scree Plot to see and help to decide how many PCs to keep.

```{r eval=FALSE, include=TRUE}
fviz_eig(pca_data)
```

**How to Interpret It:**

1. **Look for the “elbow”**

-   This is the point where the slope of the curve **levels off**.

-   It indicates that **adding more PCs beyond this point contributes little extra variance**.

-   **Keep PCs before the elbow**, discard those after.

2. **Variance Explained**

-   Each point shows **how much variance a PC explains**.

-   The **higher the point**, the more useful that PC is for summarizing the data.

Rule of Thumb:

-   If the **first few PCs** explain **70–90% of the variance**, you may choose to keep them.

## Biplot

```{r eval=FALSE, include=TRUE}
ggbiplot(pca_data)
```

What does a biplot show?

1. **Points (Observations)**

-   Each point represents a **data sample** (an observation in your dataset).

-   Their positions are based on their scores on the selected principal components (usually PC1 and PC2).

2. **Arrows (Variables)**

-   Each arrow represents a **variable (feature)** from your original data.

-   The **direction** and **length** of the arrow show how strongly that variable contributes to the PCs.

| Element | Interpretation |
|----|----|
| **Length of arrows** | Variables with longer arrows are **more strongly correlated** with the PCs shown. |
| **Angle between arrows** | Small angle -\> high positive correlation. 90° -\> no correlation. 180° -\> negative correlation. |
| **Direction of arrows** | Shows which way the variable increases. Observations in that direction have higher values for that variable. |
| **Position of points** | Samples that are close together are **similar** in terms of the principal components. |

# Example

Our goal is to predict the iris Species based on the other 4 attributes given. Subsequently we\
wish to compare our data-driven clustering to that of the true species classification. For each of\
the following six clustering methods

-   K-mean;

-   Hierarchical: Ward, Single-linkage, Complete-\
    linkage, Average-linkage, Centroid

## K-means

### Data-Preprocessing

```{r}
# omit NAs
data <- na.omit(iris)

# scale data
scaled_data <- scale(data[,-5]) #don't scale the response
```

### Optimal number of clusters

```{r}
# K-means
kmeans_fit <- kmeans(
  scaled_data,
  centers = 3,
  nstart = 25
)

# Scree-plot
fviz_nbclust(
  scaled_data,
  kmeans,
  method="wss"
)
```

3 clusters seems good

### 2D Cluster Plot

```{r}
clusplot(
  scaled_data,
  kmeans_fit$cluster,
  main = "K-means Clustering",
  color = TRUE,
  shade = TRUE,
  labels = 2,
  lines = 0
)
```

### Confusion Matrix

```{r}
table(
  kmeans_fit$cluster,
  data$Species
)
```

```{r}
#mapping
clusters <- factor(
  kmeans_fit$cluster, 
  levels = c(3,1,2), 
  labels = c("setosa","versicolor","virginica")
)

confusionMatrix(
  clusters,
  data$Species
)
```

so the confusion matrix is

50 0 0

0 39 14

0 11 36

So the accuracy is

```{r}
(39+50+36)/150
```

## Hierarchical Clustering

### Distance Matrix

```{r}
d <- dist(
  data,
  method = "euclidean"
)
```

### Ward.D

```{r}
# fit
H.fit <- hclust(
  d,
  method = "ward.D"
)
# plot
plot(H.fit)
# Add rectangles for k clusters
rect.hclust(
  H.fit, 
  k=3, 
  border = "red"
)
```

#### 2D Cluster Plot

```{r}

groups <- cutree(
  H.fit,
  k=3
)

clusplot(
  data,
  groups,
  main = "2D representation of the Cluster solution",
  color = TRUE,
  shade = TRUE,
  labels = 2,
  lines = 0
)
```

#### Confusion Matrix

```{r}
table(
  data$Species,
  groups
  
)
```

```{r}
clusters <- factor(
  groups, 
  levels =1:3, 
  labels = c("setosa","versicolor","virginica")
)

confusionMatrix(
  data$Species,
  clusters
)
```

so the confusion matrix is

50 0 0

0 50 0

0 14 36

So the accuracy is

```{r}
(50+50+36)/150
```

### Single linkage

```{r}
# fit
H.fit <- hclust(
  d,
  method = "single"
)
# plot
plot(H.fit)

# Add rectangles for k clusters
rect.hclust(
  H.fit, 
  k=3, 
  border = "red"
)
```

#### 2D Cluster Plot

```{r}

groups <- cutree(
  H.fit,
  k=3
)

clusplot(
  data,
  groups,
  main = "2D representation of the Cluster solution",
  color = TRUE,
  shade = TRUE,
  labels = 2,
  lines = 0
)
```

#### Confusion Matrix

```{r}
table(
  data$Species,
  groups
)
```

```{r}
clusters <- factor(
  groups, 
  levels =1:3, 
  labels = c("setosa","versicolor","virginica")
)

confusionMatrix(
  data$Species,
  clusters
)
```

so the confusion matrix is

50 0 0

0 50 0

0 48 2

So the accuracy is

```{r}
(50+50+2)/150
```

### Complete

```{r}
# fit
H.fit <- hclust(
  d,
  method = "complete"
)
# plot
plot(H.fit)

# Add rectangles for k clusters
rect.hclust(
  H.fit, 
  k=3, 
  border = "red"
)
```

#### 2D Cluster Plot

```{r}

groups <- cutree(
  H.fit,
  k=3
)

clusplot(
  data,
  groups,
  main = "2D representation of the Cluster solution",
  color = TRUE,
  shade = TRUE,
  labels = 2,
  lines = 0
)
```

#### Confusion Matrix

```{r}
table(
  data$Species,
  groups
)
```

```{r}
clusters <- factor(
  groups, 
  levels =1:3, 
  labels = c("setosa","versicolor","virginica")
)

confusionMatrix(
  data$Species,
  clusters
)
```

so the confusion matrix is

50 0 0

0 23 27

0 49 1

So the accuracy is

```{r}
(50+23+1)/150
```

### Average

```{r}
# fit
H.fit <- hclust(
  d,
  method = "average"
)
# plot
plot(H.fit)

# Add rectangles for k clusters
rect.hclust(
  H.fit, 
  k=3, 
  border = "red"
)
```

#### 2D Cluster Plot

```{r}

groups <- cutree(
  H.fit,
  k=3
)

clusplot(
  data,
  groups,
  main = "2D representation of the Cluster solution",
  color = TRUE,
  shade = TRUE,
  labels = 2,
  lines = 0
)
```

#### Confusion Matrix

```{r}
table(
  data$Species,
  groups
)
```

```{r}
clusters <- factor(
  groups, 
  levels =1:3, 
  labels = c("setosa","versicolor","virginica")
)

confusionMatrix(
  data$Species,
  clusters
)
```

so the confusion matrix is

50 0 0

0 50 0

0 14 36

So the accuracy is

```{r}
(50+50+36)/150
```

### Centroid

```{r}
# fit
H.fit <- hclust(
  d,
  method = "centroid"
)
# plot
plot(H.fit)

# Add rectangles for k clusters
rect.hclust(
  H.fit, 
  k=3, 
  border = "red"
)
```

#### 2D Cluster Plot

```{r}

groups <- cutree(
  H.fit,
  k=3
)

clusplot(
  data,
  groups,
  main = "2D representation of the Cluster solution",
  color = TRUE,
  shade = TRUE,
  labels = 2,
  lines = 0
)
```

#### Confusion Matrix

```{r}
table(
  data$Species,
  groups
)
```

```{r}
clusters <- factor(
  groups, 
  levels =1:3, 
  labels = c("setosa","versicolor","virginica")
)

confusionMatrix(
  data$Species,
  clusters
)
```

so the confusion matrix is

50 0 0

0 50 0

0 48 2

So the accuracy is

```{r}
(50+50+2)/150
```

### Compare K-means and Ward's

```{r}
H.fit <- hclust(
  d, 
  method="ward.D"
)

groups <- cutree(
  H.fit,
  k=3
)

ward_labels <- factor(
  groups, 
  levels = 1:3, 
  labels = c("setosa", "versicolor", "virginica")
)

kmeans_labels <- factor(
  kmeans_fit$cluster, 
  levels = c(3,1,2), 
  labels = c("setosa", "versicolor", "virginica")
)

table(
  kmeans = kmeans_labels, 
  ward = ward_labels
)
```

### Which is Better?

Ward's method is a little bit better than K-means, especially to find virginica level.

**Performance:**

-   Ward \> K-means \> Complete \> Average \> Single = Centroid

I would recommend the cluster analysis as a good way to deal with this IRIS dataset.\
K-means and complete methods are usually great, but here Ward method is the best one from the confusion matrix.

## PCA

### Compute PC's

```{r}
iris.pca <- prcomp(
  scaled_data,
  center = TRUE,
  scale.=TRUE
)
```

```{r}
summary(iris.pca)
```

**What percentage of the variations each PC would explain?**

```{r}
variance_explained <- iris.pca$sdev^2 / sum(iris.pca$sdev^2)
variance_explained * 100
```

PC1 -\> 73%

PC2 -\> 23%

PC3 -\> 3.5%

PC4 -\> 0.5%

### Biplot

```{r}
ggbiplot(iris.pca)
```

```{r}
ggbiplot(
  iris.pca,
  ellipse=TRUE,
  groups=iris$Species
)
```

### Print PC1 as Linear Combination of features

```{r}
iris.pca$rotation[,1]
```

### Result

I recommend the PCA method as an efficient way here. PCA is a very effective tool in this context.

The first two PCs account for over 95% of the variance in the original data, allowing us to visualize and classifythe species clearly.

PCA helps simplify complex data while preserving important structure and is especially useful when dealing with high-dimensional or noisy data.
