---
title: "Neural Network Example"
author: "Kachun Huang"
date: "2025-03-03"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Neural Network Classification Example

## Import Libraries

```{r}
library(neuralnet)
library(MASS)
library(tidyverse)
library(leaps)
library(car)
library(caret)
library(dplyr)
library(lmtest)
```

## Import Dataset

The banknote.csv data (see attached) were extracted from images that were taken from genuine and forged banknote-like specimens. Yes, this is a Catch Me if You Can story. For digitization, an industrial camera usually used for print inspection was used. The final images have 400 x 400 pixels. (\*3D would be called the voxels) Wavelet Transform tool were used to extract features from images.

\
There are 1,372 banknotes, and 5 variables:

1.  variance of Wavelet Transformed image (continuous)

2.  skewness of Wavelet Transformed image (continuous)

3.  curtosis of Wavelet Transformed image (continuous)

4.  entropy of image (continuous)

5.  class (binary) – this is the response variable of interest, 1 (forged) or 0 (genuine)

```{r}
bank <- read.table("/Users/thomashuang/Downloads/banknote.csv",sep=",",header = TRUE)
head(bank)
```

```{r}
str(bank)
```

## Data Pre-Processing

### Remove Missing Values

```{r}
bank <- na.omit(bank)
str(bank)
```

None of the observation have missing values.

### Factorize response

## Data Splitting

```{r}
set.seed(123)
training.samples <- bank$class %>% createDataPartition(p=0.75, list=FALSE)
train.data <- bank[training.samples,]
test.data <- bank[-training.samples,]
```

## Perceptron Model with No Hidden Layers

### with SSE Loss function

```{r}
p_model_sse <- neuralnet(
  class ~., 
  train.data, 
  hidden = 0, # no hidden layer
  err.fct = "sse", # sse loss function
  linear.output = FALSE, # non-linear output for classification
  act.fct = "logistic"# logistic activation function
  )
```

#### Plot model

```{r}
plot(p_model_sse, rep="best")
```

#### Predict on Test Data

```{r}
pred_sse <- predict(p_model_sse, test.data)
pred_sse_class <- ifelse(pred_sse > 0.5,1,0)
```

#### Confusion Matrix

```{r}
cm1 <- confusionMatrix(factor(pred_sse_class), factor(test.data$class), positive = "1")
cm1
```

When positive class = "1" Detect forged

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

#### Sensitivity

$Sensitivity= P(\text{true positive}|\text{correct}) = \frac{TP}{TP+FN}$

```{r}
181/(181+1)
```

```{r}
cm1$byClass["Sensitivity"]
```

#### Specificity

$Specificity= P(\text{true negative}|\text{correct})=\frac{TN}{TN+FP}$

```{r}
160 / (160+1)
```

```{r}
cm1$byClass["Specificity"]
```

#### Overall Accuracy

$P(\text{Predicted Correctly}) = \frac{TP+TN}{TP+TN+FP+FN}$

```{r}
(181+160)/(181+160+1+1)
```

```{r}
cm1$overall["Accuracy"]
```

### with CE Loss function

```{r}
p_model_ce <- neuralnet(
  class ~., 
  train.data, 
  hidden = 0, 
  err.fct = "ce",
  linear.output = FALSE,
  act.fct = "logistic"
  )
```

#### Plot model

```{r}
plot(p_model_ce, rep="best")
```

#### Predict on Test Data

```{r}
pred_ce <- predict(p_model_ce, test.data)
pred_ce_class <- ifelse(pred_ce> 0.5,1,0)
```

#### Confusion Matrix

```{r}
cm2 <- confusionMatrix(factor(pred_ce_class), factor(test.data$class), positive = "1")
cm2
```

When positive class = "1" Detect forged

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

#### Sensitivity

$Sensitivity= P(\text{true positive}|\text{correct}) = \frac{TP}{TP+FN}$

```{r}
179 / (179+3)
```

```{r}
cm2$byClass["Sensitivity"]
```

#### Specificity

$Specificity= P(\text{true negative}|\text{correct})=\frac{TN}{TN+FP}$

```{r}
160/(160+1)
```

```{r}
cm2$byClass["Specificity"]
```

#### Overall Accuracy

$P(\text{Predicted Correctly}) = \frac{TP+TN}{TP+TN+FP+FN}$

```{r}
(160+178)/(160+178+1+3)
```

```{r}
cm2$overall["Accuracy"]
```

## Logistic Regression

### Fit model

```{r}
logit_model <- glm(class ~ variance + skewness + curtosis + entropy, 
                   data = train.data, 
                   family = "binomial")
summary(logit_model)
```

#### Predict on Test Data

```{r}
pred_logit <- predict(logit_model, test.data)
pred_logit_class <- ifelse(pred_logit > 0.5, 1,0)
```

#### Confusion Matrix

```{r}
cm3 <- confusionMatrix(factor(pred_logit_class),factor(test.data$class), positive = "1")
cm3
```

When positive class = "1" Detect forged

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

#### Sensitivity

$Sensitivity= P(\text{true positive}|\text{correct}) = \frac{TP}{TP+FN}$

```{r}
179 / (3+179)
```

```{r}
cm3$byClass["Sensitvity"]
```

#### Specificity

$Specificity= P(\text{true negative}|\text{correct})=\frac{TN}{TN+FP}$

```{r}
160/(160+1)
```

```{r}
cm3$byClass["Specificity"]
```

#### Overall Accuracy

$P(\text{Predicted Correctly}) = \frac{TP+TN}{TP+TN+FP+FN}$

```{r}
(179+160)/(179+160+1+3)
```

```{r}
cm3$overall["Accuracy"]
```

## Perceptron Model with 1 Hidden Layers and 3 neurons

### with SSE Loss function

```{r}
p_model_layer_sse <- neuralnet(
  class ~., 
  train.data, 
  hidden = 3, 
  err.fct = "sse",
  linear.output = FALSE,
  act.fct = "logistic"
  )
```

#### Plot model

```{r}
plot(p_model_layer_sse, rep="best")
```

#### Predict on Test Data

```{r}
pred_layer_sse <- predict(p_model_layer_sse, test.data)
pred_layer_sse_class <- ifelse(pred_layer_sse > 0.5,1,0)
```

#### Confusion Matrix

```{r}
cm4 <- confusionMatrix(factor(pred_layer_sse_class), factor(test.data$class), positive = "1")
cm4
```

When positive class = "1" Detect forged

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

#### Sensitivity

$Sensitivity= P(\text{true positive}|\text{correct}) = \frac{TP}{TP+FN}$

```{r}
182/(182+0)
```

```{r}
cm4$byClass["Sensitivity"]
```

#### Specificity

$Specificity= P(\text{true negative}|\text{correct})=\frac{TN}{TN+FP}$

```{r}
161 / (0+161)
```

```{r}
cm4$byClass["Specificity"]
```

#### Overall Accuracy

$P(\text{Predicted Correctly}) = \frac{TP+TN}{TP+TN+FP+FN}$

```{r}
(161+182)/(161+182+0+0)
```

```{r}
cm4$overall["Accuracy"]
```

### with CE Loss function

```{r}
p_model_layer_ce <- neuralnet(
  class ~., 
  train.data, 
  hidden = 3, 
  err.fct = "ce",
  linear.output = FALSE,
  act.fct = "logistic"
  )
```

#### Plot model

```{r}
plot(p_model_layer_ce, rep="best")
```

#### Predict on Test Data

```{r}
pred_layer_ce <- predict(p_model_layer_ce, test.data)
pred_layer_ce_class <- ifelse(pred_layer_ce > 0.5,1,0)
```

#### Confusion Matrix

```{r}
cm5 <- confusionMatrix(factor(pred_layer_ce_class), factor(test.data$class), positive = "1")
cm5
```

When positive class = "1" Detect forged

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

#### Sensitivity

$Sensitivity= P(\text{true positive}|\text{correct}) = \frac{TP}{TP+FN}$

```{r}
182/(182+0)
```

```{r}
cm5$byClass["Sensitivity"]
```

#### Specificity

$Specificity= P(\text{true negative}|\text{correct})=\frac{TN}{TN+FP}$

```{r}
(161)/(161+0)
```

```{r}
cm5$byClass["Specificity"]
```

#### Overall Accuracy

$P(\text{Predicted Correctly}) = \frac{TP+TN}{TP+TN+FP+FN}$

```{r}
(161+182)/(161+182+0+0)
```

```{r}
cm5$overall["Accuracy"]
```

## Result

The prediction with hidden layer is better than no hidden layer in this task.
