---
title: "Neural Network in R"
author: "Kachun Huang"
date: "2025-03-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Neural Network in R

## Import library

```{r}
library(neuralnet)
```

## Fit neural network model

*syntax:*

```{r eval=FALSE, include=TRUE}
nn_model <- neuralnet(
  formula, 
  data, 
  hidden = 1, 
  stepmax = 1e+05, 
  rep = 1, 
  lifesign = "none", 
  algorithm = "rprop+", 
  err.fct = "sse", 
  linear.output = TRUE
  )

```

Parameters

+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Argument**  | **Description**                                                                                                                                                                                                                                                                                                                                                                                    |
+:=============:+:==================================================================================================================================================================================================================================================================================================================================================================================================:+
| formula       | a symbolic description of the model to be fitted.                                                                                                                                                                                                                                                                                                                                                  |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| data          | a data frame containing the variables specified in formula.                                                                                                                                                                                                                                                                                                                                        |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| hidden        | a vector of integers specifying the number of hidden neurons (vertices) in each layer.                                                                                                                                                                                                                                                                                                             |
|               |                                                                                                                                                                                                                                                                                                                                                                                                    |
|               | ex) c(4,5)                                                                                                                                                                                                                                                                                                                                                                                         |
|               |                                                                                                                                                                                                                                                                                                                                                                                                    |
|               | 4 neurons in the 1st hidden layers                                                                                                                                                                                                                                                                                                                                                                 |
|               |                                                                                                                                                                                                                                                                                                                                                                                                    |
|               | 5 neurons in the 2nd hidden layers                                                                                                                                                                                                                                                                                                                                                                 |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| err.fct       | a differentiable function that is used for the calculation of the error. Alternatively, the strings ‘sse’ and ‘ce’ which stand for the sum of squared errors and the cross-entropy can be used.                                                                                                                                                                                                    |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| linear.output | logical. If act.fct should not be applied to the output neurons set linear output to TRUE, otherwise to FALSE.                                                                                                                                                                                                                                                                                     |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| lifesign      | a string specifying how much the function will print during the calculation of the neural network. ‘none’, ‘minimal’ or ‘full’.                                                                                                                                                                                                                                                                    |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| rep           | the number of repetitions for the neural network’s training.                                                                                                                                                                                                                                                                                                                                       |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| algorithm     | a string containing the algorithm type to calculate the neural network. The following types are possible: ‘backprop’, ‘rprop+’, ‘rprop-‘, ‘sag’, or ‘slr’. ‘backprop’ refers to backpropagation, ‘rprop+’ and ‘rprop-‘ refer to the resilient backpropagation with and without weight backtracking, while ‘sag’ and ‘slr’ induce the usage of the modified globally convergent algorithm (grprop). |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| stepmax       | the maximum steps for the training of the neural network. Reaching this maximum leads to a stop of the neural network’s training process.                                                                                                                                                                                                                                                          |
+---------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

## Plot the network

*syntax:*

```{r}
plot(
  x, 
  rep = best
  )
```

| **Argument** | **Description** |
|----|----|
| x | an object of class nn |
| rep | repetition of the neural network. If rep="best", the repetition with the smallest error will be plotted. If not stated all repetitions will be plotted, each in a separate window. |

## Make prediction

```{r}
pred <- predict(nn_model, test.data)
pred_class <- ifelse(pred[,2] > 0.5,1,0)
```

## 
