---
title: "RNN and LSTM"
author: "Kachun Huang"
date: "2025-04-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = FALSE)
Sys.setenv(CLI_NUM_COLORS = 0)
```

# RNN and LSTM

## Import Library

```{r}
library(keras)
library(tensorflow)
library(reticulate)
library(tidyverse)
library(tidyquant)
library(zoo)
library(caret)
library(tidyquant)
library(magrittr)
```

## Data Pre-processing

Before building models, raw data often requires transformation. Common pre-processing steps include:

### **Normalization**

To stabilize model learning, normalize features such as stock prices using min-max scaling:

```{r eval=FALSE, include=TRUE}
data$min_lagged <- lag(data$Low)
data$max_lagged <- lag(data$High)
data$Close_norm <- (data$Close - data$min_lagged) / (data$max_lagged - data$min_lagged)
```

### **Handling Missing Values**

```{r eval=FALSE, include=TRUE}
data <- na.omit(data)
```

### Date Conversion (Optional)

```{r eval=FALSE, include=TRUE}
data$Date <- as.Date(data$Date)
```

### Train-Test Splitting

Split your data into training and test sets to evaluate your model's generalization performance:

```{r eval=FALSE, include=TRUE}
split_index <- floor(0.9 * length(series))
train_data <- series[1:split_index]
test_data <- series[(split_index + 1):length(series)]
```

-   `length(series)` returns the total number of time steps in your dataset.

-   `floor(0.9 * length(series))` calculates an index at 90% of the data.

-   `train_data` selects the first 90% as the training set.

-   `test_data` selects the remaining 10% as the test set.

### Creating Time Series Datasets

```{r eval=FALSE, include=TRUE}
create_dataset <- function(series, timesteps = 10) {
  x <- list()
  y <- c()
  for (i in 1:(length(series) - timesteps)) {
    x[[i]] <- series[i:(i + timesteps - 1)]
    y <- c(y, series[i + timesteps])
  }
  x_array <- do.call(rbind, x)
  return(list(x = x_array, y = y))
}
```

This prepares the data in a format compatible with RNNs and LSTMs.

Then apply the `create_dataset()` function:

```{r eval=FALSE, include=TRUE}
train_set <- create_dataset(train_data, timesteps)
test_set <- create_dataset(test_data, timesteps)
```

Then split `train_set` to `x_train`, `y_train` and convert it to tensor

```{r eval=FALSE, include=TRUE}
#convert to array
x_train_array <- array(as.numeric(train_set$x), dim = c(nrow(train_set$x), timesteps, 1))
y_train_array <- as.numeric(train_set$y)

# Convert to TensorFlow tensors
x_train <- tf$convert_to_tensor(x_train_array, dtype = "float32")
y_train <- tf$convert_to_tensor(y_train_array, dtype = "float32")
```

## RNN Model Building

```{r eval=FALSE, include=TRUE}
input_rnn <- layer_input(shape = c(timesteps, 1))

output_rnn <- input_rnn %>%
  layer_simple_rnn(units = 32) %>%
  layer_dense(units = 1)

model_rnn <- keras_model(inputs = input_rnn, outputs = output_rnn)
```

## LSTM Model Building

```{r eval=FALSE, include=TRUE}
input_lstm <- layer_input(shape = c(timesteps, 1))

output_lstm <- input_lstm %>%
  layer_lstm(units = 32) %>%
  layer_dense(units = 1)

model_lstm <- keras_model(inputs = input_lstm, outputs = output_lstm)
```

These models can be used for sequence-to-one prediction problems.

## **Compilation and Training**

```{r eval=FALSE, include=TRUE}
model$compile(
  optimizer = optimizer_adam(),
  loss = "mean_squared_error"
)

history <- model$fit(
  x = x_train,
  y = y_train,
  epochs = r_to_py(100L),
  batch_size = r_to_py(16L),
  verbose = 0
)
```

`r_to_py()` ensures integers are converted correctly for Python-based TensorFlow backend.

## Prediction

```{r eval=FALSE, include=TRUE}
predict_next <- function(model, input_seq_array, n_steps = 10) {
  preds <- c() # empty vector to store prediction
  current_input <- input_seq_array 
  for (i in 1:n_steps) {
    pred <- model$predict(current_input)
    preds <- c(preds, as.numeric(pred))
    updated_seq <- c(as.numeric(current_input)[-1], as.numeric(pred))
    current_input <- array(updated_seq, dim = c(1, timesteps, 1))
    current_input <- tf$convert_to_tensor(current_input, dtype = "float32")
  }
  return(preds)
}
```

This function generates `n_steps` future predictions based on a rolling window.

### Rescaling Predictions

If you used normalization, reverse it using the original min and max values:

```{r eval=FALSE, include=TRUE}
rescaled_value <- norm_value * (original_max - original_min) + original_min
```

## Model Performance

```{r eval=FALSE, include=TRUE}
mse <- mean((predicted - actual)^2)
```

## Visualizing Predictions

```{r eval=FALSE, include=TRUE}
plot(1:length(actual), actual, type = 'l', col = 'black')
lines(1:length(predicted), predicted, col = 'red')
legend("topright", legend = c("Actual", "Predicted"), col = c("black", "red"), lwd = 2)
```

# Example: sin Curve

Generate a sine wave with the function sin(x) for x ranging from 0 to 20, resulting in a total of 400 points. Plot the generated sequence.

```{r}
set.seed(123) # set seed

data <- sin(seq(0, 20, length.out = 400)) # create sin data point

# Visualize the sine wave
plot(
  as.vector(data), 
  col = 'blue', 
  type = 'l', 
  ylab = "Value", 
  main ="Sine Wave"
)
```

Divide the cleaned dataset into two parts, the **last 10 points for testing y** and **the rest for training**.

```{r}
length(data)
train_data <- data[1:(length(data)-10)]
test_data <- data[(length(data)-9):length(data)]
```

prepares the data in a format compatible with RNNs and LSTMs.

```{r}
create_dataset <- function(series, timesteps = 10) {
  x <- list()
  y <- c()
  for (i in 1:(length(series) - timesteps)) {
    x[[i]] <- series[i:(i + timesteps - 1)]
    y <- c(y, series[i + timesteps])
  }
  x_array <- do.call(rbind, x)
  return(list(x = x_array, y = y))
}
```

Create Training Set

```{r}
timesteps <- 10
train_set <- create_dataset(train_data, timesteps)
```

Convert to Tensor

```{r}
#convert to array
x_train_array <- array(as.numeric(train_set$x), dim = c(nrow(train_set$x), timesteps, 1))
y_train_array <- as.numeric(train_set$y)

# Convert to TensorFlow tensors
x_train <- tf$convert_to_tensor(x_train_array, dtype = "float32")
y_train <- tf$convert_to_tensor(y_train_array, dtype = "float32")
```

Please first build RNN with length of sequence 10 using the training data and the RNN method\
with only one RNN layer with 32 hidden units, and the loss function of ‘mse’. You will need to\
create sequences that the RNN can use to learn the pattern. Each input sequence should use 10\
sequential points to predict the next point.

```{r}
input_rnn <- layer_input(shape = c(timesteps, 1))

output_rnn <- input_rnn %>%
  layer_simple_rnn(units = 32) %>%
  layer_dense(units = 1)

model_rnn <- keras_model(inputs = input_rnn, outputs = output_rnn)

model_rnn$compile(
  optimizer = optimizer_adam(),
  loss = "mean_squared_error"
)
```

Please first build LSTM with length of sequence 10 using the training data and the LSTM method\
with only one LSTM layer with 32 hidden units, and the loss function of ‘mse’. You will need to\
create sequences that the RNN can use to learn the pattern. Each input sequence should use 10\
sequential points to predict the next point.

```{r}
input_lstm <- layer_input(shape = c(timesteps, 1))

output_lstm <- input_lstm %>%
  layer_lstm(units = 32) %>%
  layer_dense(units = 1)

model_lstm <- keras_model(inputs = input_lstm, outputs = output_lstm)

model_lstm$compile(
optimizer = optimizer_adam(),
loss = "mean_squared_error"
)
```

Plot the training loss for RNN and LSTM.

```{r}
history_rnn <- model_rnn$fit(
  x = x_train,
  y = y_train,
  epochs = r_to_py(100L),
  batch_size = r_to_py(16L),
  verbose = 0
)

history_lstm <- model_lstm$fit(
x = x_train,
y = y_train,
epochs = r_to_py(100L),
batch_size = r_to_py(16L),
verbose = 0
)
```

```{r}
plot(
  history_rnn$history$loss, 
  type = 'l', 
  col = 'red', 
  lwd = 2,
  ylim = range(c(
    history_rnn$history$loss,
    history_lstm$history$loss)
    ),
  xlab = "Epoch", 
  ylab = "Training Loss", 
  main = "Training Loss Comparison"
)

lines(
  history_lstm$history$loss, 
  col = 'blue', 
  lwd = 2
)

legend(
  "topright", 
  legend = c("RNN", "LSTM"), 
  col = c("red", "blue"),
  lwd = 2
)

```

Please compute the MSE of each model in Question 3 and Question 4 for the last 10 points.\
Compare the performance of the algorithms in Question 3 and Question 4 in length-10 forecast\
using the testing data, and plot the predictions along with the actual data.

```{r}
# Predictions
input_seq <- data[381:390]

input_seq_array <- array(
  as.numeric(input_seq), 
  dim = c(1, timesteps,1)
)

input_seq_array <- tf$convert_to_tensor(
  input_seq_array, 
  dtype = "float32"
)

predict_next <- function(model, input_seq_array, n_steps = 10) {
  preds <- c()
  current_input <- input_seq_array
  
  for (i in 1:n_steps) {
    pred <- model$predict(current_input)
    preds <- c(preds, as.numeric(pred))
    updated_seq <- c(as.numeric(current_input)[-1], as.numeric(pred))
    current_input <- array(updated_seq, dim = c(1, timesteps, 1))
    current_input <- tf$convert_to_tensor(current_input, dtype = "float32")
  }
  
  return(preds)
}
```

```{r}
# 10-step prediction
rnn_preds <- predict_next(
  model_rnn, 
  input_seq_array, 
  n_steps = 10
)
```

```{r}
lstm_preds <- predict_next(
  model_lstm, 
  input_seq_array, 
  n_steps = 10
)
```

### Comparison

```{r}
plot(
  1:400, 
  data, 
  type = 'l', 
  col = 'black', 
  lwd = 2, 
  ylab = "Value",
  xlab = "Time Step",
  main = "True vs Predicted (Last 10 Points)"
)

lines(
  391:400, 
  rnn_preds, 
  col = 'red', 
  lwd = 2
)

lines(
  391:400, 
  lstm_preds, 
  col = 'blue', 
  lwd = 2
)

points(
  391:400, 
  test_data, 
  col = 'black', 
  pch = 16
)

legend(
  "bottomleft", 
  legend = c("True", "RNN Prediction", "LSTM Prediction"),
  col = c("black", "red", "blue"), 
  lwd = 2, 
  pch = c(16, NA, NA)
)
```

# Example: AAPL

```{r}
data <- read.csv("AAPL.csv")

# transform the data type from 'chr' to 'Date'
data$Date = as.Date(data$Date)

head(data)
```

Plot the close price vs. date to visualize the data we will analyze.

```{r}
ggplot(
  data, 
  aes(x=Date, y = Close)
) + geom_line()
```

Then, use the ‘min-max scaler’ to normalize our stock price data.

Scaled_x = (x-lagged_min(x))/(lagged_max(x) – lagged_min(x)).

Please report the values of the last 10 scaled close prices.

```{r}
data$min_lagged = lag(data$Low)
data$max_lagged = lag(data$High)

data$Close_norm = (data$Close - data$min_lagged) / (data$max_lagged -data$min_lagged)
```

```{r}
model_data <- matrix(
  data$Close_norm[-1] #excluding volume
)
```

Divide the cleaned dataset into two parts, the last 10 prices for testing y and the rest for\
training (the last 10 prices in the training set will be used as testing x). Please report how many\
days of stock price are divided into the training set.

```{r}
set.seed(123)

# Create sliding window dataset
create_dataset <- function(series, timesteps = 10) {
  x <- list()
  y <- c()
  for (i in 1:(length(series) - timesteps)) {
  x[[i]] <- series[i:(i + timesteps - 1)]
  y <- c(y, series[i + timesteps])
  }
  x_array <- do.call(rbind, x)
  return(list(x = x_array, y = y))
}

# Prepare training data
timesteps <- 10

train_data <- model_data[1:(length(model_data)-10)] # training = all except last 10 days

test_data <- model_data[(length(model_data)-9):length(model_data)] # true last 10 points

train_set <- create_dataset(train_data, timesteps)

x_train_array <- array(
  as.numeric(train_set$x), 
  dim = c(nrow(train_set$x), timesteps, 1)
)

y_train_array <- as.numeric(train_set$y)

x_train <- tf$convert_to_tensor(
  x_train_array, 
  dtype = "float32"
)

y_train <- tf$convert_to_tensor(
  y_train_array, 
  dtype = "float32"
)
```

Please first build the predictive model to predict 10-days stock close price using the training\
data and the LSTM method with only one LSTM layer with 200 hidden units, and the loss\
function of ‘mse’. Please make predictions on the 10 observations in the testing set by using the\
last 10 in the training dataset and compute the Test MSE using the testing data. Scale the\
predicted stock price back and plot the 10-days predictions and the true stock close price in the\
same figure. Also, try to predict the close price of 2022-3-29.

```{r}
# Build RNN Model
input_rnn <- layer_input(shape = c(timesteps, 1))

output_rnn <- input_rnn %>%
  layer_simple_rnn(units = 32) %>%
  layer_dense(units = 1)

model_rnn <- keras_model(inputs = input_rnn, outputs = output_rnn)

model_rnn$compile(
  optimizer = optimizer_adam(),
  loss = "mean_squared_error"
)

history_rnn <- model_rnn$fit(
  x = x_train,
  y = y_train,
  epochs = r_to_py(100L),
  batch_size = r_to_py(16L),
  verbose = 0
)
```

Please first build the predictive model to predict 10-days stock close price using the training\
data and the RNN method with only one RNN layer with 200 hidden units, and the loss function\
of ‘mse’. Please make predictions on the 10 observations in the testing set by using the last 10\
in the training dataset and compute the Test MSE using the testing data. Scale the predicted\
stock price back and plot the 10-days predictions and the true stock close price in the same\
figure. Also, try to predict the close price of 2022-3-29.

```{r}
# Build LSTM Model
input_lstm <- layer_input(shape = c(timesteps, 1))

output_lstm <- input_lstm %>%
  layer_lstm(units = 32) %>%
  layer_dense(units = 1)

model_lstm <- keras_model(inputs = input_lstm, outputs = output_lstm)

model_lstm$compile(
  optimizer = optimizer_adam(),
  loss = "mean_squared_error"
)

history_lstm <- model_lstm$fit(
  x = x_train,
  y = y_train,
  epochs = r_to_py(100L),
  batch_size = r_to_py(16L),
  verbose = 0
)
```

### Comparison

Please compare the performance of the algorithms in Question 4 and Question 5 in 10-days\
forecast using the testing data

```{r}
# Prepare input for forecasting
input_seq <- model_data[(length(model_data)-10):(length(model_data)-1)] # 10 points before test

input_seq_array <- array(as.numeric(input_seq), dim = c(1, timesteps,1))

input_seq_array <- tf$convert_to_tensor(input_seq_array, dtype = "float32")

# Prediction function
predict_next <- function(model, input_seq_array, n_steps = 10) {
  preds <- c()
  current_input <- input_seq_array
  for (i in 1:n_steps) {
    pred <- model$predict(current_input)
    preds <- c(preds, as.numeric(pred))
    updated_seq <- c(as.numeric(current_input)[-1], as.numeric(pred))
    current_input <- array(updated_seq, dim = c(1, timesteps, 1))
    current_input <- tf$convert_to_tensor(current_input, dtype = "float32")
  }
  return(preds)
}
```

```{r}
rnn_preds_norm <- predict_next(
  model_rnn, 
  input_seq_array, 
  n_steps = 10
)
```

```{r}
lstm_preds_norm <- predict_next(
  model_lstm, 
  input_seq_array, 
  n_steps = 10
)
```

Scale Back

```{r}
# Initialize vectors to store scaled-back predictions and true prices
rnn_preds_original <- numeric(10)
lstm_preds_original <- numeric(10)
true_prices_original <- numeric(10)

# Scale back one by one (because each day uses its own lagged High/Low)
for (i in 1:10) {
  lagged_min <- data$Low[length(train_data) + i - 1] # lag(Low) for that day
  lagged_max <- data$High[length(train_data) + i - 1] # lag(High) for that day
  rnn_preds_original[i] <- rnn_preds_norm[i] * (lagged_max - lagged_min) + lagged_min
  lstm_preds_original[i] <- lstm_preds_norm[i] * (lagged_max - lagged_min) + lagged_min
  true_prices_original[i] <- data$Close[length(train_data) + i] # true Close at day
}
```

MSE

```{r}
mse_rnn <- mean((rnn_preds_original - true_prices_original)^2)

mse_lstm <- mean((lstm_preds_original - true_prices_original)^2)

cat("Test MSE (RNN, original scale):", mse_rnn, "\n")
cat("Test MSE (LSTM, original scale):", mse_lstm, "\n")
```

Plot 10-Day Forecast (Original Scale)

```{r}
plot(
  (length(train_data)+1):(length(train_data)+10),
  true_prices_original,
  type = 'l', 
  lwd = 2, 
  col = 'black', 
  xlab = "Time Index", 
  ylab = "Close Price (USD)",
  main = "10-Day Forecast vs True (Original Scale)"
)

lines(
  (length(train_data)+1):(length(train_data)+10),
  rnn_preds_original, 
  col = 'red', 
  lwd = 2
)

lines(
  (length(train_data)+1):(length(train_data)+10),
  lstm_preds_original, 
  col = 'blue', 
  lwd = 2
)

legend(
  "bottomleft", 
  legend = c("True Close", "RNN Prediction", "LSTM Prediction"),
  col = c("black", "red", "blue"), 
  lwd = 2
)
```

\
Plot 1-Step Ahead Forecast Only (Original Scale)

```{r}
plot(
  1, 
  true_prices_original[1], 
  type = 'p', 
  lwd = 2, 
  col = 'black',
  xlab = "Forecast Horizon (1 Day)", 
  ylab = "Close Price (USD)",
  main = "1-Day Forecast in Original Scale",
  xlim = c(0.5, 1.5),
  ylim = range(c(true_prices_original[1], rnn_preds_original[1], lstm_preds_original[1]))
)

axis(
  1, 
  at = 1, 
  labels = paste0("Day ", length(train_data)+1)
)

points(
  1, 
  rnn_preds_original[1], 
  col = 'red', 
  pch = 17
)

points(
  1, 
  lstm_preds_original[1], 
  col = 'blue', 
  pch = 17
)

legend(
  "topleft", 
  legend = c("True Close", "RNN Forecast", "LSTM Forecast"),
  col = c("black", "red", "blue"), 
  pch = c(16, 17, 17)
)
```

Model Test MSE (Original USD Scale)

-   RNN 13.13

-   LSTM 10.63

    LSTM has lower MSE than RNN. LSTM outperforms RNN when predicting actual stock prices.
