---
title: "SVM in R"
author: "Kachun Huang"
date: "2025-04-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SVM in R

## Import Library

```{r}
library(tidyverse)
library(caret)
library(kernlab)
library(e1071)
```

## Step 1: Data Pre-processing

**Feature Scaling (Standardization or Normalization)**

-   SVM is **sensitive to the scale** of features — especially with **RBF**, **polynomial**, and **linear kernels**.

**Handle Categorical Variables**

-   SVM requires **numerical input only**

-   Convert categorical features using:

    -   **One-hot encoding** (preferred)

    -   **Label encoding** (only if the categories have ordinal meaning)

**Remove or Impute Missing Values**

-   SVM does not handle `NA` or `NaN` values.

-   You must either:

    -   Remove rows with missing values

    -   Impute them (mean, median, mode, or model-based imputation)

## Step 2: Model Fitting

```{r eval=FALSE, include=TRUE}
model <- train(
  response~.,
  data = ...,
  method = "svmLinear",
  trControl = trainControl("cv", number = 10),
  preProcess = c("center","scale") #This will scale all the predicator in the model
)
```

## Step 3: Prediction

```{r eval=FALSE, include=TRUE}
predicted.classes <- model %>% predict(test.data)
```

## Step 4: Confusion Matrix

```{r eval=FALSE, include=TRUE}
confusionMatrix(
  factor(predicted.classes),
  factor(data$response),
  positive = "1"
)
```

## Hyper-parameter Tuning

We can utilize `tuneGrid` and `tuneLength` parameters in `train()` function.

### Tune C parameter

```{r eval=FALSE, include=TRUE}
model <- train(
  response~.,
  data = ...,
  method = "svmLinear",
  trControl = trainControl("cv", number = 10),
  tuneGrid = expand.grid(C = seq(0,2,length = 10))
)
```

-   This creates 20 evenly spaced values between `0` and `2`.

-   These are the values of the **regularization parameter `C`** the model will try.

### TuneLength

`caret` will automatically create a grid of  **different `C` values** (on a log scale by default), and test them via cross-validation.

```{r eval=FALSE, include=TRUE}
model <- train(
  response~.,
  data = ...,
  method = "svmLinear",
  trControl = trainControl("cv", number = 10),
  tuneLength = n
)
```

# Example

## Import Data

```{r}
data <- read.csv("Titanic.csv")
```

## Data Pre-Processing

### Remove Columns

```{r}
data <- subset(
   data,
   select = -c(PassengerId,Name,Ticket,Cabin)
)
```

### Remove Missing Value in Age Column

```{r}
data <- subset(data, is.na(Age) == FALSE)
```

### Factorize Categorical Variable

```{r}
data$Pclass <- as.factor(data$Pclass)
data$Survived <- as.factor(data$Survived)
```

### Standardize Numeric Column

```{r}
data$Age <- scale(data$Age)
data$Fare <- scale(data$Fare)
```

### Data Splitting

```{r}
set.seed(123)
training.samples <- data$Survived %>%
createDataPartition(p = 0.75, list = FALSE)
train.data <- data[training.samples, ]
test.data <- data[-training.samples, ]
str(train.data) 
```

```{r}
str(test.data)
```

## Model Fitting

```{r}
set.seed(123)
model <- train(
  Survived ~., data = train.data, method = "svmLinear",
  trControl = trainControl("cv", number = 10),
  # preProcess = c("center","scale")
  # This will scale all the predictors in the model
)
```

### Prediction

```{r}
predicted.classes <- model %>% predict(test.data)
```

### Confusion Matrix

```{r}
confusionMatrix(
  factor(predicted.classes), 
  factor(test.data$Survived), 
  positive = '1'
)
```

## Hyper-parameter Tuning

### with tuneGrid = expand.grid(C = seq(0, 2, length = 20))

```{r}
set.seed(123)
model <- train(
  Survived ~., data = train.data, method = "svmLinear",
  trControl = trainControl("cv", number = 10),
  # preProcess = c("center","scale")
  # This will scale all the predictors in the model
  tuneGrid = expand.grid(C = seq(0, 2, length = 20))
)
```

#### Plot

```{r}
plot(model)
```

#### Best Tune

```{r}
model$bestTune
```

Here you can directly use the fitted model or you can refit a model using the best tuning parameter.

#### Model with best Tune

```{r}
set.seed(123)
model <- train(
  Survived ~., data = train.data, method = "svmLinear",
  trControl = trainControl("cv", number = 10),
  tuneGrid = expand.grid(C = model$bestTune)
)

predicted.classes <- model %>% predict(test.data)

confusionMatrix(factor(predicted.classes), factor(test.data$Survived), positive = '1')
```

### With TuneLength = 10, radial basis kernel

```{r}
set.seed(123)
model <- train(
  Survived ~., 
  data = train.data, 
  method = "svmRadial",
  trControl = trainControl("cv", number = 10),
  # preProcess = c("center","scale")
  # This will scale all the predictors in the model
  tuneLength = 10
)
```

#### Best Tune

```{r}
model$bestTune
```

#### Prediction

```{r}
predicted.classes <- model %>% predict(test.data)
```

#### Confusion Matrix

```{r}
confusionMatrix(
  factor(predicted.classes), 
  factor(test.data$Survived), 
  positive = '1'
)
```

### TuneLength = 4, polynomial basis kernel

```{r}
set.seed(123)
model <- train(
  Survived ~., 
  data = train.data, 
  method = "svmPoly",
  trControl = trainControl("cv", number = 10),
  # preProcess = c("center","scale")
  # This will scale all the predictors in the model
  tuneLength = 4
)
```

#### Best Tune

```{r}
model$bestTune
```

#### Prediction

```{r}
predicted.classes <- model %>% predict(test.data)
```

#### Confusion Matrix

```{r}
confusionMatrix(
  factor(predicted.classes), 
  factor(test.data$Survived), 
  positive = '1'
)
```

# Result

\
**Overall accuracy rate:**\
Linear: 0.7978\
Radial basis kernel: 0.8146 (the best)\
Polynomial kernel: 0.809
